\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}

\label{ch:introduction}
Heat exchangers are key devices in industrial and energy systems that often need precise temperature control. Their slow dynamics, operational constraints, and nonlinear behavior make control challenging, especially for tzpical industrial methods. These factors drive the adoption of advanced control strategies.

Model Predictive Control (MPC) has become a standard approach for controlling constrained systems in process engineering. 
Its theoretical foundations are well established in the literature, where MPC is formulated as a finite-horizon optimization problem that explicitly incorporates system constraints \cite{rawlings_mpc}. 
This framework enables systematic control design for both Single-Input Single-Output (SISO) and Multi-Input Multi-Output (MIMO) systems, making MPC suitable for a wide range of industrial applications, including thermal processes.


MPC has been applied to heat exchanger control in both industrial and pilot-scale settings. 
Reported applications include outlet temperature regulation under constrained conditions, where MPC outperformed PID control by reducing temperature variance by 20--40\%, while maintaining stable operation near input and output limits \cite{mpc_heat_exchanger}. These outcomes have motivated further research into improved modeling techniques for MPC, including data-driven approaches that better capture nonlinear system dynamics.


Classical linear models remain widely used in industrial applications due
to their simplicity and ease of identification. In thermal systems such as
heat exchangers, step-response experiments are commonly used to identify
low-order linear models. These models typically describe the dominant
dynamics using a steady-state gain and a single time constant, which is
often sufficient for control design in a limited operating region
\cite{ljung_sysid}. Low-order linear models are particularly common in SISO heat exchanger
configurations \cite{mpc_heat_exchanger}.


Although such models perform well near a nominal operating point, their accuracy decreases in the presence of nonlinearities. This limitation motivates the adoption of more advanced modeling approaches for predictive control \cite{morari1999} and has led to increased research interest in data-driven and nonlinear modeling techniques.

Data-driven modeling provides an alternative to classical first-principles and step-response-based approaches.  Rather than relying on physical laws or simplified assumptions, data-driven methods construct system representations directly from measured input--output data. 
This approach captures complex and nonlinear system behavior more accurately, particularly when physical modeling is challenging or when the system operates over a broad range of conditions. 
\cite{ljung_sysid}.

Data-driven modeling provides an alternative to classical first-principles and step-response-based approaches. Rather than relying only on physical laws or simplified assumptions, data-driven methods construct system representations directly from measured input--output data. This approach can be useful for thermal systems, where nonlinear heat transfer, changing operating conditions, and unmeasured disturbances make just physical modeling difficult \cite{ljung_sysid}.

Several recent works show the use of data-driven models. Brandes investigated data-driven modeling of heat pumps and thermal storage units for MPC and compared static models with dynamic state-space approaches based on Dynamic Mode Decomposition with control (DMDc) and Sparse Identification of Nonlinear Dynamics with control (SINDyc). In that work, the models were evaluated for heating and cooling case studies, including their prediction performance and model complexity. The study also discussed neural-network-based models, noting that they can represent nonlinear behavior but usually require more training data and have lower interpretability compared with simpler state-space approaches \cite{brandes2022}.

Another example is the work of Wu et al., where support vector regression (SVR) was used for data-driven prediction of steady-state conditions in heat exchangers. SVR is based on the support vector machine (SVM) principle, but it is used for regression problems instead of classification. In this approach, the model learns the relation between measured operating variables and the steady-state output without requiring a detailed analytical description of the heat exchanger. The authors applied the method to preheaters and steam generators and showed that the SVR-based model could predict steady-state operating conditions with mean errors below 3.5~\%, while also maintaining robustness under input fluctuations \cite{wu2025svr}.



Among these approaches, the Koopman operator framework establishes a theoretical basis for representing nonlinear dynamical systems using linear operators acting on lifted state spaces. The concept was introduced by Koopman \cite{koopman1931}, and later expanded by Mezi\'c, who demonstrated its practical application in analyzing nonlinear system behavior \cite{mezic2005}.

A key aspect of the Koopman framework is the lifting function 
$g(\cdot)$, which maps the original system states or measurements into a higher-dimensional space. In this lifted space, nonlinear dynamics can be represented by a linear prediction model. The selection or construction of the lifting function is therefore critical to the accuracy of the resulting Koopman model.


Various strategies for constructing the lifting function $g(\cdot)$ have been proposed in the literature. Early data-driven approaches utilized predefined dictionaries of basis functions, such as polynomials, radial basis functions, or Fourier modes. Extended Dynamic Mode Decomposition (EDMD) is a well-known example, approximating the Koopman operator using linear regression in a specified function space \cite{williams2015}. The effectiveness of such methds depends on the selected dictionary and can be constrained for complex or high-dimensional systems.

Recent research has increasingly focused on learning the lifting function directly from data. Brunton, Proctor, and Kutz~\cite{brunton2016} demonstrated that sparse data-driven representations can identify system dynamics in the absence of explicit physical models. Building on this foundation, Korda and Mezi\'c~\cite{korda2018} introduced Koopman-based linear predictors for nonlinear systems and established their applicability in optimal control problems.
Neural-network-based lifting functions have further  advanced this approach by providing flexible representations of nonlinear dynamics while retaining a linear structure suitable for control design \cite{koopman_mpc_review}.


The Koopman framework has also been incorporated into MPC frameworks \cite{korda2018}.
In Koopman-based MPC, the learned linear model is employed directly within the MPC optimization problem, allowing nonlinear system behaviors to be represented using a linear prediction model. 
This integration enables direct comparison of Koopman-based MPC and classical linear MPC under identical constraints and cost functions. This thesis ivestigates how Koopman-based modelling influences control performance.


Applications of Koopman-based MPC to thermal systems have been demonstrated in several recent studies. For example,  Koopman–Model Predictive Control was implemented in a warm-water supply system, where a data-driven linear Koopman model regulated outlet temperature under variable demand and operating conditions, resulting in improved tracking and constraint handling compared to conventional control approaches \cite{miyashita2022koopman}.
Related research has also addressed vehicle thermal management, where Koopman-based MPC was used for temperature regulation of automotive thermal subsystems with nonlinear heat transfer dynamics, while respecting actuator and safety constraints \cite{chen2025koopman_tms}.





